---
title: "GGplot"
author: "Cristina Cheroni"
date: "04 January 2017"
output: html_document
---

The purpose of this document is to illustrate the main functions of ggplot2 package. I will consider both main plot types and format options that could be applied to multiple types. 


## PLOT TYPES

### SCATTERPLOT

**Arguments and Examples at [http://ggplot2.tidyverse.org/reference/geom_point.html](this link)**

**First example:** scatterplot using data from Weekly dataset (ISLR package)

geom_point is the main command for scatterplot generation. 
I also use options for specifying x and y labels and plot title

```{r ScatterplotI, fig.align='center', fig.width=10, fig.height=7}
library(ggplot2)
library(ISLR)

ggplot(Weekly, aes(Year, Volume)) +
    geom_point() +
    xlab('Year') + ylab('Volume') +
    ggtitle('Weekly dataset: \n relationship between Volume and Year')
```

**To change color for all the points:**
```{r ScatterplotII, fig.align='center', fig.width=10, fig.height=7}
ggplot(Weekly, aes(Year, Volume)) +
    geom_point(color='blue') +
    xlab('Year') + ylab('Volume') +
    ggtitle('Weekly dataset: \n relationship between Volume and Year')
```

**For conditional color and size:**
```{r ScatterplotIII, fig.align='center', fig.width=10, fig.height=7}
ggplot(Weekly, aes(Year, Volume)) +
    geom_point(aes(col=Direction, size=Lag1)) +
    xlab('Year') + ylab('Volume') +
    ggtitle('Weekly dataset: \n relationship between Volume and Year')
```
This can be applied to both continuous and categorical data. 


**To add horizontal or vertical lines: geom_vline and geom_hline**
```{r ScatterplotIV, fig.align='center', fig.width=10, fig.height=7}
ggplot(Weekly, aes(Year, Volume)) +
    geom_point(aes(col=Year)) +
    geom_vline(xintercept=2005, col='blue') +
    geom_hline(yintercept=5, col='blue') +
    xlab('Year') + ylab('Volume') +
    ggtitle('Weekly dataset: \n relationship between Volume and Year')
```


**To add an interpolating line:** 
```{r ScatterplotV, fig.align='center', fig.width=10, fig.height=7}
ggplot(Weekly, aes(Year, Volume)) +
    geom_point(aes(col=Year)) +
    geom_smooth(method='lm') +
    xlab('Year') + ylab('Volume') +
    ggtitle('Weekly dataset: \n relationship between Volume and Year')
```

For other methods to be used with geom_smooth see [http://ggplot2.tidyverse.org/reference/geom_smooth.html](this page)


**To connect the observations we can use geom_line() or geom_path()**. In another example:  


```{r ScatterplotVI, fig.align='center', fig.width=10, fig.height=7}
data <- data.frame('Treatment'=c(rep('A', 3), rep('B', 3)), 'Dose'=c(0.5, 1, 2, 0.5, 1, 2), 'Response'=c(10, 20, 25, 5, 7, 9))
ggplot(data, aes(x=Dose, y=Response)) +
    geom_point(aes(col=Treatment)) +
    geom_line(aes(col=Treatment)) 
```

```{r ScatterplotVII, fig.align='center', fig.width=10, fig.height=7}
data <- data.frame('Treatment'=c(rep('A', 3), rep('B', 3)), 'Dose'=c(0.5, 1, 2, 0.5, 1, 2), 'Response'=c(10, 20, 25, 5, 7, 9))
ggplot(data, aes(x=Dose, y=Response)) +
    geom_point(aes(col=Treatment)) +
    geom_step(aes(col=Treatment)) 
```



### HISTOGRAMS

**First example: plotting histograms for a single variable**
I rely on Auto dataset (ISLR library) and represent the acceleration value.

```{r HistogramI, fig.align='center', fig.width=9, fig.height=6}
ggplot(Auto, aes(acceleration)) + 
geom_histogram()
```

*Second example: plotting histograms for a two variables*
For Auto dataset, I represent the acceleration and cylinders. The representation on a single plot of the these two variables is not really meaningful, but it is just an example. 

```{r HistogramII, fig.align='center', fig.width=9, fig.height=6}
ggplot(Auto, aes(acceleration)) + 
geom_histogram(fill='#4876FF75', binwidth=0.5) +
geom_histogram(aes(cylinders), fill='#FF450075', binwidth=0.5)    
```

*Third example: differential colouring according to a variable*

```{r HistogramIII, fig.align='center', fig.width=9, fig.height=6}
ggplot(Auto, aes(acceleration, fill=factor(origin))) + 
geom_histogram(alpha=0.5, binwidth=0.5)    
```

Origin is a numerical variable in the original dataset, so I have to specify to treat it as a factor. 


### DENSITY PLOT

**First example: random generated data**

```{r DensityData, fig.align='center', fig.width=9, fig.height=6} 
k1 = rnorm(20, mean= -1.25, sd=sqrt(1))
k2 = rnorm(20, mean= 1.25, sd=sqrt(1))
dataset <- data.frame(k1, k2)
```

Simple plot for one distribution:
```{r DensityI, fig.align='center', fig.width=9, fig.height=6} 
ggplot(dataset, aes(k1)) + 
geom_density()
```
Changing bandwidth with adjust and setting a fill color
```{r DensityII, fig.align='center', fig.width=9, fig.height=6} 
ggplot(dataset, aes(k1)) + 
geom_density(adjust=0.5, fill='#4876FF')
```

Plot with overlay of two distributions and a vertical line
```{r DensityIII, fig.align='center', fig.width=9, fig.height=6} 
ggplot(dataset, aes(k1)) + 
geom_density(adjust=0.5, fill='#4876FF75') + 
geom_density(aes(k2), adjust=0.5, fill='#FF450075') +
# to add a vertical line
geom_vline(xintercept=0, linetype = "longdash", colour='grey25')
```

**Second example on Auto dataset: differential colouring according to one variable**

On Auto dataset from ISLR library, I represent the weight of the veichle considering the number of cylinders. 
Since in the original dataset the number of cylinder is a numerical variable, I have to transform it in a factor. 

```{r DensityIV, fig.align='center', fig.width=9, fig.height=6}
ggplot(Auto, aes(weight, fill=factor(cylinders))) +
geom_density(alpha=0.5)
```


### BOXPLOT

Information at the link: http://docs.ggplot2.org/current/geom_boxplot.html

*First example: simple boxplot*

Considering the Auto dataset, I plot horsepower stratifed for cylinders (transformed in to a factor)

```{r BoxI, fig.align='center', fig.width=9, fig.height=6}
ggplot(Auto, aes(as.factor(cylinders), horsepower)) +
geom_boxplot()
```

We can change fill color and border color and set the title and axis labels. 

```{r BoxII, fig.align='center', fig.width=9, fig.height=6}
ggplot(Auto, aes(as.factor(cylinders), horsepower)) +
    geom_boxplot(fill='palegreen', color=c('grey100', 'grey75', 'grey45', 'grey20', 'black')) +
    xlab('Number of Cylinders') + ylab('HorsePower') +
    ggtitle ('Horsepower stratified for cylinders')
```

**Third more complex plot:**  

 * with varwidt we set the width of the boxes according to the number of observations in each class. 
 * notch = TRUE makes notched boxplots. Notches are used to compare groups; if the notches of two boxes do not overlap, this suggests that the medians are significantly different.  
 * coord_flip to set the box horizontally

```{r BoxIII, fig.align='center', fig.width=9, fig.height=6}
ggplot(Auto, aes(as.factor(cylinders), horsepower)) +
geom_boxplot(fill='palegreen', color='black', outlier.color='darkred', varwidth=TRUE, notch=TRUE) +
    xlab('Number of Cylinders') + ylab('HorsePower') +
    ggtitle ('Horsepower stratified for cylinders') +
    coord_flip()
```


We can also visualize the single observations **adding geom_jitter**: 

```{r BoxIV, fig.align='center', fig.width=9, fig.height=6}
ggplot(Auto, aes(as.factor(cylinders), horsepower)) +
    geom_boxplot(fill='palegreen', color='black', outlier.color='darkred') +
    geom_jitter(color='darkgreen', width=0.6, shape=18) +
    xlab('Number of Cylinders') + ylab('HorsePower') +
    ggtitle ('Horsepower stratified for cylinders')
```


### VIOLIN PLOTS

Violin plots are similar to boxplots. However, they are more informative because, instead of displaying quantiles, they represent the kernel density distribution of the data sets. 

More information at [http://ggplot2.tidyverse.org/reference/geom_violin.html](this link)*First example: simple violin plot*

Considering the Auto dataset, I plot horsepower stratifed for cylinders (transformed in to a factor)

```{r ViolinI, fig.align='center', fig.width=9, fig.height=6}
ggplot(Auto, aes(as.factor(cylinders), horsepower)) +
geom_violin()
```

Changing the color and inserting titles: 
```{r ViolinII, fig.align='center', fig.width=9, fig.height=6}
ggplot(Auto, aes(as.factor(cylinders), horsepower)) +
    geom_violin(fill='palegreen') +
    xlab('Number of Cylinders') + ylab('HorsePower') +
    ggtitle ('Horsepower stratified for cylinders')
```

Setting the area of the plot proprortional to the number of observations and displaying quantiles. **N.B**: these are the quantiles of the density estimate, therefore could be different from the quantiles of the distribution as shown by a box plot.  
```{r ViolinIII, fig.align='center', fig.width=9, fig.height=6}
ggplot(Auto, aes(as.factor(cylinders), horsepower)) +
    geom_violin(fill='palegreen', scale='count', draw_quantiles=c(0.25, 0.5, 0.75)) +
    xlab('Number of Cylinders') + ylab('HorsePower') +
    ggtitle ('Horsepower stratified for cylinders')
```

We can represent on the same plot the violin and the boxplot

```{r ViolinIV, fig.align='center', fig.width=9, fig.height=6}
ggplot(Auto, aes(as.factor(cylinders), horsepower)) +
    geom_violin(fill='palegreen', draw_quantiles=c(0.25, 0.5, 0.75)) +
    geom_boxplot(width=0.1) +
    xlab('Number of Cylinders') + ylab('HorsePower') +
    ggtitle ('Horsepower stratified for cylinders')
```

We can also visualize the sigle observations:

```{r ViolinV}
ggplot(Auto, aes(as.factor(cylinders), horsepower)) +
    geom_violin(col='green', draw_quantiles=c(0.25, 0.5, 0.75)) +
    geom_jitter(color='forestgreen', width=0.5, size=0.75) +
    xlab('Number of Cylinders') + ylab('HorsePower') +
    ggtitle ('Horsepower stratified for cylinders')
```


Da sistemare sotto: vedere se si può sostituire jitter con violin che mostra meglio i punti con la forma del violin 
```{r ViolinVI}
#ggplot(Auto, aes(as.factor(cylinders), horsepower)) +
#    geom_violin(col='darkgreen', draw_quantiles=c(0.25, 0.5, 0.75)) +
#    geom_dotplot(color='forestgreen',binaxis = "x", stackdir = "center") +
#    xlab('Number of Cylinders') + ylab('HorsePower') +
#    ggtitle ('Horsepower stratified for cylinders')
```

### BARPLOT

Information at: 
https://rpubs.com/escott8908/RGC_Ch3_Gar_Graphs
http://docs.ggplot2.org/current/geom_bar.html


**To represent the number of observations for each category**
In this way we represent the counts of observation in each category. 
```{r BarI}
ggplot(Auto, aes(cylinders)) +
geom_bar()
```

Da rivedere
```{r BarII}
ggplot(Auto, aes(cylinders)) +
geom_bar() + 
    geom_bar(aes(weight=displacement))
```



# horizontal back-to-back barplot 
ggplot (data = Numbers, aes(x=Sample, y=Genes, fill=Modulation)) +
    geom_bar(data=subset(Numbers, Modulation=='Up-Regulated'), stat='identity') + 
    geom_bar(data=subset(Numbers, Modulation=='Down-Regulated'), stat='identity', position='identity', mapping=aes(y=-Genes)) +
    scale_y_continuous(breaks=seq(-300, 300, 25), labels = abs (seq(-300, 300, 25)), limits=c(-310, 310)) +
    coord_flip() + 
    scale_fill_brewer(palette = rev("Set1")) +
    theme_bw()



## COLORS, FILL AND TRANSPARENCY

**Example using boxplot**

```{r ColI, fig.align='center', fig.width=9, fig.height=6}
ggplot(OJ, aes(Purchase, PriceDiff)) +
geom_boxplot()
```

To change the colour of the box:
```{r ColII, fig.align='center', fig.width=9, fig.height=6}
ggplot(OJ, aes(Purchase, PriceDiff)) +
geom_boxplot(col='blue')
```


To change the fill: 
```{r ColIII, fig.align='center', fig.width=9, fig.height=6}
ggplot(OJ, aes(Purchase, PriceDiff)) +
geom_boxplot(fill='lightblue1')
```

To change the fill: 
```{r ColVI, fig.align='center', fig.width=9, fig.height=6}
ggplot(OJ, aes(Purchase, PriceDiff)) +
geom_boxplot(fill='lightblue1', col='blue')
```

The same can be applied to geom_bar, geom_violin, geom_line, etc. 


**Example using scatterplots**: College dataset, we examine the relationship between the number of applications received and the number of applications accepted. 

```{r ColBI, fig.align='center', fig.width=6, fig.height=6}
ggplot(College, aes(x=Apps, y=Accept)) +
geom_point()
```

Add conditional colouring according to a continuus variable (% of faculty with a PhD):
```{r ColBII, fig.align='center', fig.width=6, fig.height=6}
ggplot(College, aes(x=Apps, y=Accept, col=PhD)) +
geom_point(alpha=0.5))
``


Add conditional colouring according to a categorical variable (Private or Public college):
```{r ColBII, fig.align='center', fig.width=6, fig.height=6}
ggplot(College, aes(x=Apps, y=Accept, col=Private)) +
geom_point()
``





scale colour

* scale_colour_gradient
* scale_colour_manual
    scale_colour_manual(values=colorRampPalette(c('cornflowerblue', 'purple'))(12))
* scale_colour_hue # default colors - intensity and croma can be modified by adjusting l and c parameters - h to select the specrum of colour
* scale_colour_

http://www.sthda.com/english/wiki/ggplot2-colors-how-to-change-colors-automatically-and-manually



# change x and y axis breaks, set limits, change to log: 
scale_x_continuous
scale_y_continuous

scale_y_continuous(breaks=c(seq(0,60, by=10)), limits=c(0, 60))

http://docs.ggplot2.org/current/scale_continuous.html

# Set axis limits directly without scale call: 
+ xlim(0,10)
+ ylim(-20, 20)

facet

to change text in the titles of the facets: 
theme(strip.text.x = element_text(size = 8, colour = "darkblue"))

legend: 
legend.position and legend text (es. for size) in theme

http://www.cookbook-r.com/Graphs/Legends_(ggplot2)/




geom_smooth: 

method can be used to specify how to generate the fitting: lm, gml, ... Default is loess if dataset < 1000, gam if > 10000
formula can be also used 
span is used to adjust the amount of smoothing; higher value for less local adjustment
To fit a logistic regression model:
geom_smooth(method='glm', method.args = list(family = "binomial"), color='orchid')
To fit a linear model: 
geom_smooth(method='lm')



Importante per conversione ggplot - plotly
https://plot.ly/ggplot2/user-guide/
https://plot.ly/r/reference/#scatter











# to change the order of the categories:

scale_x_discrete(limits = c())


# === LEGENDS

http://www.cookbook-r.com/Graphs/Legends_(ggplot2)/

to change ledend position: legend.position='bottom' in theme





# We can add the points using also geom_jitter
ggplot(AutoBin, aes(mpg01, horsepower)) +
geom_boxplot(fill='white', colour=c('forestgreen', 'darkred')) + 
geom_jitter (width=0.2)

if we eant to see label, we can apply jitter to geom_text



### 
